RELIKE: Reionization Effective Likelihood from Planck 2018 Data

### 2021-04-16 paper/rei.tex
Updates from Overleaf

\begin{figure}
\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green). With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.  \todo{add SRollv2. convention: put labels only when we do comparisons}}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]
{paper/plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian.pdf}
\caption{Cumulative optical depth in the two step model (as in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).  Agreement between the exact, KDE PC and Gaussian PC likelihoods is again excellent and show that a small amount of high-$z$ ionization is allowed. The form of the model requires that the cumulative optical depth steadily declines above the first step at $z_1 \sim 6-7$ especially for $z\gtrsim 15$.
}
\label{fig:plot_taugtz_two_step_contours}
\end{figure}

\todo{change z1 z2 to zre and zhi everywhere.}
The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
 \bea
x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{1})-y(z)}{\Delta y} \right] \right\} \notag \\
&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_highz}
 \eea
where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z_1)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{1})$ just as in the first example for a sharper step.
We choose the second step at $z_{2}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{1} \lesssim z \lesssim z_2$. We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$. We show examples
of this two-step model in Fig.~\ref{fig:two_step_model}.

In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.  They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data. [double check this too.]
Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given a total optical depth $\tau$. 
For both the full likelihood and effective likelihood runs, we adopt the flat priors on $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$. 
%except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model.

@@

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
\caption{Cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ from the model-independent PC approach with and without the physicality prior.   The prior helps exclude models with negative contributions at high-$z$ but conservatively allows some negative values at $z \sim z_{\rm max}=30$ where the oscillations in the PCs no longer cancel due to the cumulative integration.  Upper limits are nearly unaffected. \todo{change all tau cumulative plots for PC to $\tau_{\rm PC}$;} \todo{add (with prior)}} 
}
% no prior: 95\% C.L.  = [-0.01935756790924691, 0.021029019067727237]
% with prior: {'z': 15, 'mean': 0.00694422689213187, 'std': 0.007568921831067864, 'limit68': [-0.0007215571045808687, 0.014684826477572082], 'limit95': [-0.006316156744039184, 0.02284538046347211]}
%
%}
\label{fig:plot_taugtz_2018_with_vs_without_physicality_prior}
\end{figure}

@@

In the future, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f. adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects. \ch{[Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data? More importantly with data in general that wants to vary As during sampling? Put prior on As consistent with Planck?]}.

@@

[Any comments on Ahn et al. \url{https://arxiv.org/pdf/2011.03582.pdf} and \url{https://arxiv.org/pdf/2006.16828.pdf}?]

>>>>

\begin{figure}
\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green). With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]
{paper/plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian.pdf}
\caption{Cumulative optical depth in the two step model (as in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).  Agreement between the exact, KDE PC and Gaussian PC likelihoods is again excellent and show that a small amount of high-$z$ ionization is allowed. The form of the model requires that the cumulative optical depth steadily declines above the first step at $z_{\rm re} \sim 6-7$ especially for $z\gtrsim 15$.
}
\label{fig:plot_taugtz_two_step_contours}
\end{figure}

The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
 \bea
x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm hi}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_highz}
 \eea
where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z_{\rm re})^{1/2}\Delta z_{\rm re}$ where $\Delta z_{\rm re} = 0.015\, (1+z_{\rm re})$ just as in the first example for a sharper step.
We choose the second step at $z_{\rm hi}=28$ with $\Delta z_{\rm hi} = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_{\rm hi}$. We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$. We show examples
of this two-step model in Fig.~\ref{fig:two_step_model}.

In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.  They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_{\rm re}$ to be conservative on the preference of $\tauhi$ in the data. 
Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given a total optical depth $\tau$. 
For both the full likelihood and effective likelihood runs, we adopt the flat priors on $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$. 
%except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model.

@@

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
\caption{Cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ from the model-independent PC approach with and without the physicality prior.   The prior helps exclude models with negative contributions at high-$z$ but conservatively allows some negative values at $z \sim z_{\rm max}=30$ where the oscillations in the PCs no longer cancel due to the cumulative integration.  Upper limits are nearly unaffected. 
}
% no prior: 95\% C.L.  = [-0.01935756790924691, 0.021029019067727237]
% with prior: {'z': 15, 'mean': 0.00694422689213187, 'std': 0.007568921831067864, 'limit68': [-0.0007215571045808687, 0.014684826477572082], 'limit95': [-0.006316156744039184, 0.02284538046347211]}
%
%}
\label{fig:plot_taugtz_2018_with_vs_without_physicality_prior}
\end{figure}

@@

In the future, we expect our code release RELIKE to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f. adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects. \ch{[Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data? More importantly with data in general that wants to vary As during sampling? Put prior on As consistent with Planck?]}.

@@

\todo{[Any comments on Ahn et al. \url{https://arxiv.org/pdf/2011.03582.pdf} and \url{https://arxiv.org/pdf/2006.16828.pdf}?]
}
